Revealing Benefits, Costs, and Tradeoffs of Spatial Management in Marine Systems Arrising from Coupled Economic and Ecological Dynamics

### 2022-12-21 03_marlin_paper.qmd
out for internal review

---
title: "Policy Implications of Coupled Economic and Ecological Dynamics in Aquatic Systems"
author:
    - Daniel Ovando
    - Darcy Bradley
    - Echelle Burns
    - Lennon Thomas
    - James Thorson
date: "`r Sys.Date()`"
format:
  pdf:
    toc: true
    fig-pos: 'H'
  html: default
  docx: default
execute:
  echo: false
  message: false
  warning: false
  cache: false
bibliography: references.bib
params:
  run_name: ["v0.3"]
abstract: "Communities around the world are increasingly looking to spatial strategies for managing marine ecosystems. However, the design of spatial management strategies faces numerous challenges. Marine ecosystems are highly dynamic and opaque, and extractive entities such as fishing fleets respond endogenously to changes in marine ecosystems in ways that depend upon ecological and policy context. We present a framework for efficient simulation of the bio-economic dynamics of marine systems, designed to support both management and research. We demonstrate the model's capabilities by focusing on two case studies on the conservation and food security impacts of marine protected areas (MPAs). We highlight how coupled bio-economic dynamics can result in complex trade-offs between conservation, fisheries yields, and distributional outcomes. "
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
---

@@

sci_to_com <- taxize::sci2comm(unique(marlin_inputs$scientific_name)) %>% 
  bind_rows(.id = "critter") %>% 
  pivot_longer(everything(),names_to = "critter",values_to = "common_name") %>% 
  mutate(common_name = stringr::str_to_title(common_name))

@@

@@

Communities around the world are increasingly looking to spatial
strategies for managing marine ecosystems. For example, the "30x30"
movement calls for implementation of marine protected areas, a form of
spatial management, across 30% of the world's oceans by the year 2030.
While the objectives of individual spatial management strategies vary
widely, these types of policies are often supported based on their
ability to provide protection, along with potential broader benefits, to
swaths of marine ecosystems with one management action, with potentially
fewer costs and complications than for example assessing, setting, and
enforcing catch quotas for individual species.

However, the design of spatial management strategies faces numerous
challenges. Marine ecosystems are highly dynamic and opaque, and
extractive entities such as fishing fleets respond to changes in marine
ecosystems in highly complex ways. These dynamics pose challenges even
for single-species assessment and management strategies [@field2006a],
which are only amplified when we consider management policies
specifically designed around spatial features in the oceans. For
example, policies such as MPAs designed for both conservation and food
security must consider factors such as the optimal size and placement of
a protected area network given a wide array of life history, species
distributions, exploitation levels, fleet dynamics, and policy-dependent
behavior [@reimer2017], all of which may change over time, particularly
given the impacts of climate change.

These challenges are amplified by the fact that our knowledge regarding
the succesful design and implementation of spatial management strategies
such as MPAs is largely based on theoretical studies; the size
[@ovando2021a] and time required [@nickols2019] to see substantial
effects from many MPAs means that we have relatively little empirical
evidence as to their large-scale effects across a range of
social-ecological settings [@mcclanahan2021].

@@

To help address this challenge, we present a novel bio-economic modeling
framework called [`marlin`](https://danovando.github.io/marlin/) that
allows for efficient simulation of spatio-temporal biological and
economic dynamics. `marlin` allows users to simulate the impacts
fisheries management policies across a range of species targeted by
various fleets across heterogeneous and dynamic seascapes. This model
can help users seek Parteo-optimal solutions that produce win-wins or
minimize tradeoffs across multiple objectives, such as food security and
conservation [@lester2013].

Here we present the core methods and functionality of `marlin`, and
demonstrate its use in two case-study applications; a coral reef system
and a pelagic system. Our results show how interacting ecological,
economic, and design attributes can affect the degree of benefits,
costs, and tradeoffs between conservation and food security outcomes of
MPAs. More broadly, this paper both demonstrates the functionality of
our model and the critical need for considering sufficiently realistic
coupled ecological and economic dynamics in policy evaluations.

@@

As a demonstration of the use of this modeling framework, we explore two
general situations

1.  Trade-offs and distributional outcomes for food security and
    ecosystem biomass in a coral-reef fishery
2.  Implications of climate-driven range shifts for MPA design in a
    pelagic fishery

@@

Below we provide a summary of the `marlin` model, as well as details of
the generalized simulation exercises and case study analyses. We note
that `marlin` is a complex software package with many options to control
detailed behaviors. Given that these specific features will evolve over
time, we do not describe in detail every option of the model here.
Rather, we focus on the equations and methods describing the core
foundations of our framework for efficiently simulating bio-economic
dynamics in aquatic systems.

@@

`marlin` simulates the dynamics of one or more species, currently best
representing fish-like species, using age-structured population
dynamics. Each time step, species move throughout the simulated seascape
using both diffusion and "taxis" (active movement towards preferred
habitat), experiences natural and potentially fishing mortality, and if
in a spawning season spawn using one of the possible forms of density
dependence implemented in the model, with potentially auto-correlated
deviations from deterministic reproductive dynamics.

These species are caught by fishing fleets. A fishing fleet is defined
in `marlin` by a set of fishers that can have different fishing skill,
prices, and selectivity for individual species (each of which we denote
as a *metier*). For example, both a longline and purse seine fleet may
capture bigeye (*Thunnus obesus*) and skipjack (*Katswamus pelamus*)
tunas, but the longline fleet can be made much more likely to catch
bigeye than skipjack, and *vice versa*. Each fishing fleet distributes
its fishing effort in space according to a specified spatial allocation
rule, for example in proportion to profit-per-unit-effort, conditional
on the presence of any spatial restrictions such as MPAs.

@@

The dynamics of each fleet can be modified by management in a variety of
ways. Managers can impose size limits for individual species within each
fleet. They can also impose total catch quotas for each species in each
time step. When quotas are activated, if the total catch across all
fleets for a given species under the model's effort dynamics would
exceed the allowable quota for that species, the total effort for each
fleet is decrease proportional to its contribution to the total catch
until the quota is satisfied. As an alternative to catch quotas,
managers can set an effort cap per fleet, which prevents effort from
exceeding a given amount under open-access dynamics, though effort may
be reduced below this cap if required by the profit equation (i.e. the
fleet can choose to fish less than the quota or effort cap). `marlin`
also allows users to specify closed fishing seasons for one or more
species in the system. During closed seasons, no effort is allocated to
patches that would result in capture of the covered species, potentially
resulting in zero effort if the species cannot be avoided. Under these
conditions, the amount of effort in the next open season is a function
of the effort and profits in the last open fishing season. Lastly,
managers can specify locations of no-take MPAs, which could in theory
change in size and location. When MPAs are implemented, fishing effort
that used to operate inside the MPAs can either leave the fishery, or be
redistributed outside the the MPA (the default behavior).

Users can use the model to explore general properties of spatially
explicit social-ecological systems, or can parameterize it to explore
key uncertainties of more tactical applications. As an example of the
later case, we provide two case studies here designed to reflect the
kinds of questions stakeholders might face if designing an MPA network
for a coral reef or pelagic system.

@@

Each fishing fleet (*f*) generates catches or yield (*Y*), revenues
(*R*), costs (*C*) from individual species (*s*) that it targets in a
given time step (*t*) and patch (*p*). The fleet then makes decisions
around how much and where to fish, subject to regulatory constraints,
based on th total profitability across all species in space and time.

@@

Total catches or yield *Y* are calculated through the Baranov equation
[@baranov1918], which appropriately accounts for the total instantaneous
mortality (both fishing and natural) and divides the total amount of the
population biomass (*b*) killed between natural (*m*) and fishing (*u*)
sources, the amount of biomass killed through fishing being assigned to
yield *Y*.

@@

This means that each species experiences a total mortality at age in a
time step in a patch, individual fractions of which are portioned off as
catches and subsequently revenues for each fleet. Given revenues *R*, we
then calculate profits $\phi$ based on revenues and costs.

@@

Costs *C* are calculated as a function of $\gamma_f$, that the scales
the cost per unit effort for fleet *f*,$\beta$ allows for the cost of
effort to scale non-linearly, and $\eta$ is an additional cost per unit
effort of fishing in each patch *p*, allowing for the model to account
for travel costs for different patches. Travel costs are calculated
based on the Euclidean distance of each patch to the nearest port of a
given fleet, users can specify any number from zero to the number of
patches of port locations. When no ports are specified, travel costs are
zero.

@@

The more complex option is "open access". Under open access, the total
amount of effort in a given time step for fleet *f* is a function of the
profitability of that fleet in the proceeding time steps in which
fishing was open, where $\theta$ controls the responsiveness of fleet
*f* to the log of the ratio of total revenues *R* to total costs *C*,
and approximates the percentage amount effort changes in response to a
one unit changes in the log revenue to cost ratio.

@@

Revenue based spatial distribution is not likely to be very realistic;
in general we would expect fishing fleets to respond to profits on some
core level. However, due to the complexity of parameterizing cost
functions fleet dynamics are often evaluated based on yield or revenue
alone, and so we include those scenarios here to allow users to evalaute
the potential implications of this choice. The decision on whether to
allocate the fleet based on absolute or relative (per unit effort)
metrics is more complex. When effort represents the actions of separate
and individual fishing actors (e.g. independent fishing vessels), a
per-unit-effort strategy may be more realistic, in which fishers
distribute themselves based on the expected catch of their individual
efforts [@hilborn1987]. Conversely, a system defined by a sole owner
seeking to maximize total profits might be better represented by a fleet
model based on total profits

@@

The underlying population model used is an age structured single-species
model in the manner of @ovando2021a. The population model requires many
parameters. However, if the user supplies either a scientific or common
name for the species in question (scientific preferred), the model will
supply default values for that species based on the values reported in
FishLife [@thorson2020]. All of the life history values can be replaced
with user-supplied values as needed though, the defaults simply ensure
that reasonable and internally consistent values for the species in
question are supplied as a starting point.

@@

`marlin`'s movement dynamics are based on a continuous-time Markov chain
(CTMC), as described in @thorson2021a. Within this framework, the model
allows for movement to be broken down into three components of
*diffusion*, *advection* (follow currents), and *taxis* (active movement
towards say a preferred habitat). For now, our model focuses just on the
diffusion and taxis components of this model, though future extensions
could easily incorporate advection vectors from oceanographic models. In
this way then, `marlin` allows users to run anything from a simple
Gaussian dispersal kernel up to a system governed by species that
passively diffuse around a dynamic thermal range.

We briefly the general CTMC method here, but see @thorson2021a for a
detailed description of the method. Under this framework, movement from
each patch to each other patch in the system in a given timestep *t* for
life stage *a* of species *s* is defined by a movement matrix
$\boldsymbol{M}_{t,s,a}$. $\boldsymbol{M}_{t,s,a}$ is calculated as the
matrix exponent of the sum of the diffusion and taxis matrices scaled by
the width of the time step that the model is running under.

@@

Where $\boldsymbol{D}$ is the diffusion matrix and $\boldsymbol{X}$ is
the taxis matrix.The diffusion matrix is a function of a user specified
diffusion rate, see @thorson2021a.

@@

While this CTMC approach to movement simulation is to date not commonly
seen in the marine modeling literature, it has numerous advantages that
warrant its seeming complexity. First, when only diffusion is present,
the model will generalize to the familiar dynamics of a Gaussian
dispersal kernel. However, the time component *t* in @eq-movement allows
the movement model to be run on arbitrarily fine time steps. Second, the
*taxis* model allow for more active habitat choices by species, allowing
us to simulate both preferences of species in space and time
efficiently. Lastly, this form has the advantage that the parameters are
directly estimable from real data. So, if provided with say spatial
abundance data and a tagging study, users can estimate the movement
parameters in the same manner as @thorson2021a, and then pass their
estimated parameters to `marlin` for simulation.

@@

Maturity is then calculated as a logistic function where $l_{mat}$ is
the length at which on average 50% of individuals are sexual maturity,
and $\delta_{mat}$ is the units of length beyond $l_{mat}$ at which on
average 95% of fish are sexually mature.

@@

Fisheries management is often concerned with measuring stock status
relative to maximum sustainable yield (MSY) based reference points,
though the exact level of stock status desired by societal objectives
may vary widely. MSY based reference points present a problem for a
multi-fleet and spatial-temporal model such as `marlin`. MSY and the
fishing mortality rate that would produce MSY, $F_{MSY}$ are a function
of fishery selectivity. Fishery selectivity in this model can vary by
fleet, and species can be distributed unevenly in both space and time.
This means that the net effective fishing selectivity on a species can
vary depending on the dynamics at a given moment, making the definition
of an equilibrium concept such as MSY challenging.

As such, we do not report MSY based reference points in the model by
default, with the exception of the spawning stock biomass at MSY
($SSB_{MSY}$) which is more conserved across selectivity patterns as it
is primarily a function of the mortality and flow of recruits relative
to the maturity ogive. There are many different strategies for
estimating reference points conditional on selectivity and regimes, see
citations xx. We leave it to users to define and find relevant reference
point as required by their specific needs.

@@

We include two examples demonstrating how `marlin` can be used to
support marine spatial planning. In the first, we show how `marlin` can
be used to compare the total and distributional impacts of marine
protected areas design in a heavily fished coral reef ecosystem. In the
second, we demonstrate how `marlin` can be used to assess components of
climate resilience of alternative MPA design strategies.

Each of the case studies contains too many parameters and options to be
succinctly presented in the text here, and we are concerned that
attempting to duplicate all of the parameters of the model here would
simply introduce error. As such, readers should consult the accompanying
code (xx provide clear instructions for which files xx) to view the
precise details of each simulation. Targeted tactical applications
though should be sure to consider and document all decisions made around
model parameters.

@@

Across all of our case studies we make use of four potential "rule of
thumb" MPA design strategies:

1.  Rate: MPAs are placed in based on the pre-MPA depletion-weighted
    catch relative to the total catch in a patch. So, patches with high
    rates of catch of depleted species relative to total catch are
    prioritized.

2.  Avoid_fishing: MPAs locations are prioritized inversely to fishery
    catches. So, patches with low catches are prioritized over patches
    with high catches.

3.  Target_fishing: MPAs locations are prioritized proportionally to
    fishery catches. So, patches with high total catches are prioritized
    over patches with low catches.

4.  Area: For the "area" strategy, we simply assign patches in
    sequential order until the desired total MPA size is reached.

MPAs never to our knowledge randomly placed; their locations are decided
based on stakeholder negotiation based on available information and
objectives. In theory the design MPA networks can be optimized through
the use of a modeling framework, and indeed conditional on the validity
of the model itself this process is likely to produce better outcomes
than manually-designed strategies [@rassweiler2012; @rassweiler2014].
However, in practice the ability of communities to conduct numerical
optimization of MPA networks can be constrained by the number of
objectives and dimensionality of the operating model, or by simple
capacity of local stakeholders. As such, we show here how `marlin` can
be used to examine the performance of potentially sub-optimal but more
broadly usable design strategies. We allow all MPAs to be designed in a
"mosaic" fashion, but users can easily extend the analysis to for
example compare outcomes between "contiguous" and "mosaic" MPA designs
[@pons2022].

@@

These species are targeted by two different fleets. Fleet one primarily
targets the grouper and nearshore snapper populations, while obtaining
some funds from sharks. Fleet one has a dome shaped selectivity pattern
for both grouper and shallow-reef snapper due to demand for "plate
sized" fish (citation), and logistic selectivity for sharks, as any
shark caught is retained. This fleet is a subsistence fleet, where the
local community takes every possible opportunity to fish that they can,
and as such we model it as a "constant effort" fishery.

The second fleet primarily targets the deep-water snapper population.
While plate-sized snapper are still prized from this population, we
model fleet two's selectivity as logistic due to high levels of discard
mortality for deep-water snapper resulting from barotrauma. Fleet two
catches some grouper, though less than fleet one, and receives no price
for sharks due to the requirements of a certification program through
which they sell their deep-water snapper, though they do accidentally
catch and kill them on occasion. This fleet operates under open-access
dynamics, as fishing is not the primary means of subsistence and so
short-term effort expands and contracts in respond to profitability of
the primarily grouper-driven fishery.

Each fleet has a different home-port location, and incurs an additional
cost for fishing each patch in the system in proportion to the euclidean
distance from each patch to each fleet's home port.

We simulated the outcomes for both food security and conservation for
each of the species and both of the fleets as a function of both MPA
size and MPA design strategy. For this exercise, MPAs are placed with
perfect information and have no design constraints for continuity (i.e.
the selected MPA design can be as patchwork as dictated by the data).

@@

We modeled our pelagic use case loosely on the characteristics of the
Western and Central Pacific Ocean tuna fisheries. We simulate
trajectories of `r n_distinct(names(blue_fauna))` species commonly
caught in the region, including both the highly abundance skipjack tuna
(*Katsuwonus pelamis*) to the heavily depleted oceanic whitetip shark
(*Carcharhinus longimanus*) (@fig-blue-ssb). We pulled publicly
available data on catch-per-unit-effort of each of these species from
the WCPO as a proxy for baseline habitat distributions (@fig-blue).

>>>>

---
title: "Revealing Benefits, Costs, and Tradeoffs of Spatial Management in Marine Systems Arrising from Coupled Economic and Ecological Dynamics"
author:
    - Daniel Ovando
    - Darcy Bradley
    - Echelle Burns
    - Lennon Thomas
    - James Thorson
date: "Updated `r Sys.Date()`"
format:
  pdf:
    toc: true
    fig-pos: 'H'
  html: default
  docx: default
execute:
  echo: false
  message: false
  warning: false
  cache: false
bibliography: references.bib
params:
  run_name: ["v0.3"]
abstract: "Abstract: Designing effective spatial management strategies is challenging because marine ecosystems are highly dynamic and opaque, and extractive entities such as fishing fleets respond endogenously to ecosystem changes in ways that depend upon ecological and policy context. We present a tool, `marlin`, that can be used to efficiently simulate the bio-economic dynamics of marine systems in support of both management and research. We demonstrate `marlin’s` capabilities by focusing on two case studies on the conservation and food production impacts of marine protected areas (MPAs): a coastal coral reef and a pelagic tuna fishery. In the coastal coral reef example, we explore how heterogeneity in species distributions and fleet preferences can affect distributional outcomes of MPAs. In the pelagic case study, we show how our model can be used to assess the climate resilience of different MPA design strategies, as well as the climate sensitivity of different fishing fleets. This paper demonstrates how intermediate complexity simulation of coupled bio-economic dynamics can help communities predict and potentially manage trade-offs between conservation, fisheries yields, and distributional outcomes of management policies affected by spatial bio-economic dynamics."
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
---

@@

sci_to_com <- taxize::sci2comm(unique(marlin_inputs$scientific_name),simplify= TRUE,db='ncbi') %>% 
  bind_rows(.id = "critter") %>% 
  pivot_longer(everything(),names_to = "critter",values_to = "common_name") %>% 
  mutate(common_name = stringr::str_to_title(common_name))

@@

Abstract: Designing effective spatial management strategies is challenging because marine ecosystems are highly dynamic and opaque, and extractive entities such as fishing fleets respond endogenously to ecosystem changes in ways that depend upon ecological and policy context. We present a tool – marlin -- to efficiently simulate the bio-economic dynamics of marine systems, designed to support both management and research. We demonstrate `marlin’s` capabilities by focusing on two case studies on the conservation and food production impacts of marine protected areas (MPAs): a coastal coral reef and a pelagic tuna fishery. In the coastal coral reef example, we explore how heterogeneity in species distributions and fleet preferences can affect distributional outcomes of MPAs. In the pelagic case study, we demonstrate how our model can be used to assess the climate resilience of different MPA design strategies, as well as the climate sensitivity of different fishing fleets. This paper demonstrates how intermediate complexity simulation of coupled bio-economic dynamics can help communities predict and potentially manage trade-offs between conservation, fisheries yields, and distributional outcomes resulting from spatial management.

@@

Communities around the world are increasingly looking to spatial
strategies for managing marine ecosystems . For example, the "30x30"
movement calls for implementation of marine protected areas (MPAs), a
form of spatial management, across 30% of the world's oceans by the year
2030 [@grorud-colvert2021 and references therein]. However , designing
spatial management strategies to achieve desired objectives -- which may
include recovery of overfished species, increased food production and
economic well being, and equitable distribution of benefits - is not
straightforward. Different species have different resilience to fishing,
distributions in space and time, and value to fishing fleets; fishing
fleets themselves can have varying ranges of incentives and reactions to
policy structures. These dynamics pose challenges even for
single-species assessment and management strategies [@field2006a], which
are only amplified when we consider management policies specifically
designed around spatial features in the oceans.

For example, policies such as MPAs designed for both conservation and
food production must consider factors such as the optimal size and
placement of a protected area network given a wide array of life
history, species distributions, exploitation levels, fleet dynamics, and
policy-dependent behavior [@reimer2017], all of which may change over
time, particularly given the impacts of climate change. Efforts to
effectively design spatial management strategies such as MPAs are
further constrained by a lack of empirical evidence describing the size
[@ovando2021a] and time required [@nickols2019] for MPAs to produce
substantial effects across a range of social-ecological settings
[@mcclanahan2021].

@@

To help address this challenge, we present a novel bio-economic modeling
tool called [`marlin`](https://danovando.github.io/marlin/) that allows
for efficient simulation of spatio-temporal biological and economic
dynamics. `marlin` allows users to simulate the impacts of marine
management policies across a range of species targeted by various fleets
across heterogeneous and dynamic seascapes. This model can help users
seek Parteo-optimal solutions that produce win-wins or minimize
tradeoffs across multiple management objectives, such as food production
and conservation [@lester2013].

Here we present the core methods and functionality of `marlin`, and
demonstrate its use in two case-study applications; a coral reef system
and a pelagic system. Our results show how interacting ecological,
economic, and design attributes can affect the degree of benefits,
costs, and tradeoffs between conservation and food production outcomes
of MPAs. More broadly, this paper both demonstrates the functionality of
our model and the critical need for considering sufficiently realistic
coupled ecological and economic dynamics in policy evaluations.

@@

As a demonstration of the use of this modeling two, we explore two
general situations

1.  Trade-offs and distributional outcomes for food production and
    multi-species biomass in a coral-reef fishery
2.  Implications of climate-driven range shifts for MPA design in a
    pelagic fishery

@@

Below we provide a summary of the `marlin` model, as well as details of
the generalized simulation exercises and case study analyses. We note
that `marlin` is a complex software package with many options to control
detailed behaviors. Given that these specific features will evolve over
time, we do not describe in detail every option of the model here.
Rather, we focus on the equations and methods describing the core
foundations of our tool for efficiently simulating bio-economic dynamics
in aquatic systems.

@@

`marlin` simulates the dynamics of one or more species, currently best
representing fish-like species, using age-structured population
dynamics. Each time step, species move throughout the simulated seascape
using both diffusion and "taxis" (active movement towards preferred
habitat), experiences natural and potentially fishing mortality, and if
in a spawning season spawn using one of the possible forms of density
dependence implemented in the model, with the potential to include
auto-correlated stochastic deviations in the amount of offspring
produced, generally called "recruits".

These species are caught by fishing fleets. A fishing fleet is defined
in `marlin` by a set of fishers that can have different fishing skill,
prices, and selectivity for individual species (each of which we denote
as a *metier*). For example, both a longline and purse seine fleet may
capture bigeye (*Thunnus obesus*) and skipjack tunas, but the longline
fleet can be made much more likely to catch bigeye than skipjack, and
*vice versa*. Each fishing fleet distributes its fishing effort in space
according to a specified spatial allocation rule, for example in
proportion to profit-per-unit-effort, conditional on the presence of any
spatial restrictions such as MPAs.

@@

The dynamics of each fleet can be modified by management in a variety of
ways. Managers can impose size limits for individual species within each
fleet. They can also impose total catch quotas for each species in each
time step. When quotas are activated, if the total catch across all
fleets for a given species under the model's effort dynamics would
exceed the allowable quota for that species, the total effort for each
fleet is decreased proportional to its contribution to the total catch
until the quota is satisfied. As an alternative to catch quotas,
managers can set an effort cap per fleet, which prevents effort from
exceeding a given amount under open-access dynamics, though effort may
be reduced below this cap if required by the profit equation (i.e. the
fleet can choose to fish less than the quota or effort cap). `marlin`
also allows users to specify closed fishing seasons for one or more
species in the system. During closed seasons, no effort is allocated to
patches that would result in capture of the covered species, potentially
resulting in zero effort if the species cannot be avoided. Under these
conditions, the amount of effort in the next open season is a function
of the effort and profits in the last open fishing season. Lastly,
managers can specify locations of no-take MPAs, which could in theory
change in size and location. When MPAs are implemented, fishing effort
that used to operate inside the MPAs can either leave the fishery, or be
redistributed outside the MPA (the default behavior).

Users can use the model to explore general properties of spatially
explicit social-ecological systems, or can parameterize it to explore
key uncertainties of more tactical applications. As an example of the
later case, we provide two case studies here designed to reflect the
kinds of questions stakeholders might face if designing an MPA network
for a nearshore coral reef or a pelagic system.

@@

Each fishing fleet (*f*) generates catches or yield (*Y*), revenues
(*R*), costs (*C*) from individual species (*s*) that it targets in a
given time step (*t*) and patch (*p*). The fleet then makes decisions
around how much and where to fish, subject to regulatory constraints,
based on the total profitability across all species in space and time.

@@

Total catches or yield *Y* are calculated through the Baranov equation
[@baranov1918], which appropriately accounts for the total instantaneous
mortality (both fishing and natural) and divides the total amount of the
population biomass (*b*) killed between natural (*m*) and fishing (*u*)
sources, with the amount of biomass killed through fishing assigned to
yield *Y*.

@@

This means that each species experiences a total mortality at age in a
time step in a patch, individual fractions of which are portioned off as
catches and subsequently revenues for each fleet. Given revenues *R*, we
then calculate profits $\phi$ based on revenues and costs. Costs *C* are
calculated as a function of $\gamma_f$ that the scales the total cost
per unit effort for fleet *f*. $\beta$ allows for the cost per unit
effort effort to scale non-linearly, and $\eta$ is an additional cost
per unit effort of fishing in each patch *p*, allowing for the model to
account for travel costs for different patches. Travel costs are
calculated based on the Euclidean distance of each patch to the nearest
port of a given fleet, and users can specify any number from zero to the
number of patches of port locations. When no ports are specified, travel
costs are zero.

@@

@@

The more complex option is "open access". Under open access, the total
amount of effort in a given time step for fleet *f* is a function of the
profitability of that fleet in the proceeding time steps in which
fishing was open, where $\theta$ controls the responsiveness of fleet
*f* to the log of the ratio of total revenues *R* to total costs *C*,
and approximates the proportional change in effort in response to a one
unit changes in the log revenue to cost ratio.

@@

Revenue based spatial distribution is not likely to be very realistic;
in general we would expect fishing fleets to respond to profits on some
core level. However, due to the complexity of parameterizing cost
functions fleet dynamics are often evaluated based on yield or revenue
alone, and so we include those scenarios here to allow users to evaluate
the potential implications of this choice. The decision on whether to
allocate the fleet based on absolute or relative (per unit effort)
metrics is more complex. When effort represents the actions of separate
and individual fishing actors (e.g. independent fishing vessels), a
per-unit-effort strategy, in which fishers distribute themselves based
on the expected catch of their individual efforts, may be more realistic
[@hilborn1987]. Conversely, a system defined by a sole owner seeking to
maximize total profits might be better represented by a fleet model
based on total profits

@@

The underlying population model used is an age structured single-species
model in the manner of @ovando2021a. The population model requires many
parameters. However, if the user supplies either a scientific or common
name for the species in question (scientific preferred), the model will
supply default values for that species based on the values reported in
FishLife [@thorson2020]. All of the life history values can
alternatively be replaced with user-supplied values as needed; the
default values simply ensure that reasonable and internally consistent
values for the species in question are supplied as a starting point.

@@

`marlin`'s movement dynamics are based on a continuous-time Markov chain
(CTMC), as described in @thorson2021a. Within this framework, the model
allows for movement to be broken down into three components of
*diffusion*, *advection* (follow currents), and *taxis* (active movement
towards say a preferred habitat). For now, our model focuses just on the
diffusion and taxis components of this model, though future extensions
could easily incorporate advection vectors from oceanographic models. In
this way, `marlin` allows users to run anything from a simple Gaussian
dispersal kernel up to a system governed by species that passively
diffuse around a dynamic thermal range.

We provide a brief overview of the the general CTMC method here, see
@thorson2021a for a detailed description. Under this framework, movement
from each patch to each other patch in the system in a given timestep
*t* for life stage *a* of species *s* is defined by a movement matrix
$\boldsymbol{M}_{t,s,a}$. $\boldsymbol{M}_{t,s,a}$ is calculated as the
matrix exponent of the sum of the diffusion and taxis matrices scaled by
the width of the time step that the model is running under.

@@

Where $\boldsymbol{D}$ is the diffusion matrix and $\boldsymbol{X}$ is
the taxis matrix. The diffusion matrix is a function of a user specified
diffusion rate [@thorson2021a].

@@

While this CTMC approach to movement simulation is to date not commonly
seen in the marine modeling literature, it has numerous advantages that
warrant its seeming complexity. First, when only diffusion is present,
the model will generalize to the familiar dynamics of a Gaussian
dispersal kernel. However, the time component *t* in @eq-movement allows
the movement model to be run on arbitrarily fine time steps. Second, the
taxis model allow for more active habitat choices by species, allowing
us to simulate both preferences of species in space and time
efficiently. Lastly, this form has the advantage that the parameters are
directly estimable from real data. So, if provided with say spatial
abundance data and a tagging study, users can estimate the movement
parameters in the same manner as @thorson2021a, and then pass their
estimated parameters to `marlin` for simulation.

@@

Maturity is then calculated as a logistic function where $l_{mat}$ is
the length at which on average 50% of individuals are sexually mature,
and $\delta_{mat}$ is the unit of length beyond $l_{mat}$ at which on
average 95% of fish are sexually mature.

@@

Fisheries management is often concerned with measuring stock status
relative to maximum sustainable yield (MSY) based reference points,
though the exact level of stock status desired by management or societal
objectives may vary widely. MSY based reference points present a problem
for a multi-fleet and spatial-temporal model such as `marlin`. MSY and
the fishing mortality rate that would produce MSY, $F_{MSY}$ are a
function of fishery selectivity. Fishery selectivity in this model can
vary by fleet, and species can be distributed unevenly in both space and
time. This means that the net effective fishing selectivity on a species
can vary depending on the dynamics at a given moment, making the
definition of an equilibrium concept such as MSY challenging.

As such, we do not report MSY based reference points in the model by
default, with the exception of the spawning stock biomass at MSY
($SSB_{MSY}$) which tends to be conserved across selectivity patterns as
it is primarily a function of the mortality and flow of recruits
relative to the maturity ogive. There are many different strategies for
estimating reference points in spatially explicit systems [@kapur2021a].
We leave it to users to define and find relevant reference point as
required by their specific needs.

@@

We include two examples demonstrating how `marlin` can be used to
support marine spatial planning. In the first, we show how `marlin` can
be used to compare the total and distributional impacts of MPAs designed
in a heavily fished coral reef ecosystem. In the second, we demonstrate
how `marlin` can be used to assess components of climate resilience of
alternative MPA design strategies in a pelagic system.

Each of the case studies contains too many parameters and options to be
succinctly presented in the text here, and we are concerned that
attempting to duplicate all of the parameters of the model here would
simply introduce error. As such, readers should consult the accompanying
code (xx provide clear instructions for which files xx) to view the
precise details of each simulation. Targeted tactical applications must
carefully consider and document all decisions made around model
parameters.

@@

Across all of our case studies we make use of four potential "rule of
thumb" MPA design strategies (xx add citation here) :

1.  Depletion Rate (*rate*): MPAs are placed based on the pre-MPA
    depletion-weighted catch relative to the total catch in a patch. So,
    patches with high rates of catch of depleted species relative to
    total catch are prioritized.

2.  Avoid_fishing (*avoid_fishing*): MPAs locations are prioritized
    inversely to fishery catches. So, patches with low catches are
    prioritized over patches with high catches.

3.  Target_fishing (*target_fishing*): MPAs locations are prioritized
    proportionally to fishery catches. So, patches with high total
    catches are prioritized over patches with low catches.

4.  Area target (*area*): For the "area" strategy, we simply assign
    patches in sequential order until the desired total MPA size is
    reached.

To our knowledge, MPAs are never randomly placed; their locations are
decided based on stakeholder negotiation given available information and
objectives. In theory, the design of MPA networks can be optimized
through the use of a modeling framework, and indeed conditional on the
validity of the model itself this process is likely to produce better
outcomes than manually-designed strategies [@rassweiler2012;
@rassweiler2014]. However, designing optimal MPA networks becomes
increasingly difficult as the range of objectives and the complexity of
the model increase. Therefore, while `marlin` provides some support for
[optimization](https://danovando.github.io/marlin/articles/optimize-mpas.html),
we focus here on the design and performance of these more "rule of
thumb" design strategies that may be more accessible to a wider range of
users. We allow all MPAs to be designed in a "mosaic" fashion, but users
can easily extend the analysis to for example compare outcomes between
"contiguous" and "mosaic" MPA designs [@pons2022].

@@

These species are targeted by two different fleets. Fleet one primarily
targets the grouper and nearshore snapper populations, but will land any
incidentally captured sharks. Fleet one has a dome shaped selectivity
pattern for both grouper and shallow-reef snapper due to demand for
"plate sized" fish [@kindsvater2017], and logistic selectivity for
sharks, as any shark caught is retained. This fleet is a subsistence
fleet, where the local community takes every possible opportunity to
fish that they can, and as such we model it as a "constant effort"
fishery.

The second fleet primarily targets the deep-water snapper population.
While plate-sized snapper are still prized from this population, we
model fleet two's selectivity as logistic due to high levels of discard
mortality for deep-water snapper resulting from barotrauma. Fleet two
catches some grouper, though less than fleet one, and receives no price
for sharks due to the requirements of a certification program through
which they sell their deep-water snapper, though accidental captures do
occur, which results in mortality. This fleet operates under open-access
dynamics, as fishing is not the primary means of subsistence and so
short-term effort expands and contracts in response to profitability of
the primarily grouper-driven fishery.

Each fleet has a different home-port location, and incurs an additional
cost for fishing each patch in the system in proportion to the Euclidean
distance from each patch to each fleet's home port.

We use `marlin` to simulate the outcomes for both food production and
conservation for each of the species and both of the fleets as a
function of both MPA size and MPA design strategy. For this exercise,
MPAs are placed with perfect information and have no design constraints
for continuity (i.e. the selected MPA design can be as patchwork as
dictated by the data).

@@

We model our pelagic use case loosely on the characteristics of the
Western and Central Pacific Ocean tuna fisheries. We simulate
trajectories of `r n_distinct(names(blue_fauna))` species commonly
caught in the region, including both the highly abundant skipjack tuna
to the heavily depleted oceanic whitetip shark (*Carcharhinus
longimanus*) (@fig-blue-ssb). We use publicly available data on
catch-per-unit-effort of each of these species from the WCPO as a proxy
for baseline habitat distributions, noting that where possible
fishery-independent abundance indices would be preferable (@fig-blue).

These pelagic species are caught by a longline fleet that primarily
targets large adult tunas such as bigeye and yellowfin for high-grade
consumption, and a purse-seine fishery that primarily targets skipjack
tunas for bulk canning. Selectivities were modeled as logistic for the
longline fleet, and dome-shaped for the purse-seine fleet, with
selectivity values set to broadly match the patterns reported in
relevant [stock
assessments](https://oceanfish.spc.int/en/ofpsection/sam/sam/210-stock-assessment-input-and-result-files-)
for each species. Both fleets operate under open-access dynamics with an
effort cap. The effort cap was set at the level of effort that resulted
in the desired levels of depletion for each species under open-access
dynamics (@fig-blue), intended to simulate a scenario where managers
step in to prevent further expansion of fishing effort in a fully
developed fishery. For forward-simulation, open-access dynamics can
result in effort decreasing in response to profitability, but cannot
result in effort beyond the effort cap set for each fleet.
